Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-22 intro.tex
minor formatting to intro.tex

As a corollary of the undecidability of the halting problem
\cite{halting-problem}, however, no single forecasting schema is ideal
for all noise-free deterministic signals\cite{weigend-book}---let
alone all real-world time-series data sets.  This leads naturally to
an important and challenging question: given a complicated real-valued
time series, does there exist a forecast model that can utilize the
information (if any) that is being transmitted from past to future by
the underlying generating process?  A first step in answering this
question is to reliably quantify where on the complexity spectrum a
given time series falls; a second step is to determine how complexity
and predictability are related.  These are the goals of this paper.

@@

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation\cite{crutchfield2003}.  
\label{page:redundancy}
Our approach
exploits this decomposition in order to assess how much predictive
structure is present in a signal---i.e., where it falls on the
complexity spectrum mentioned above.  We define \emph{complexity} as a
particular approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.
That is, we view a random-walk time series (which exhibits high
entropy) as purely complex, whereas a periodic signal, which exhibits
low entropy, is on the low end of the complexity spectrum.  We argue
that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy through
ordinal analysis---is an effective way to assess the complexity of a
given time series.  Permutation entropy is ideal for our purposes
because it works with real-valued data and is known to converge to the true entropy value. Other
existing techniques either require specific knowledge of the
generating process or produce biased values of the
entropy~\cite{bollt2001}.

@@

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  While this is not a common laboratory
experiment, it is a highly appropriate choice here.  Computers are
among the most complex engineered artifacts in current use.  Modern
microprocessor chips contain multiple processing units and multi-layer
memories, for instance, and they use complicated hardware/software
strategies to move data and threads of computation across those
resources.  As a result, the processor and memory loads during the
execution of even a very simple program can evolve in a very
complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance trace
of a three-line loop that repeatedly initializes the upper triangle of
a matrix in column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the program can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate the
  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

>>>>

As a corollary of the undecidability of the halting
problem~\cite{halting-problem}, however, no single forecasting schema
is ideal for all noise-free deterministic
signals~\cite{weigend-book}---let alone all real-world time-series
data sets.  This leads naturally to an important and challenging
question: given a complicated real-valued time series, does there
exist a forecast model that can utilize the information (if any) that
is being transmitted from past to future by the underlying generating
process?  A first step in answering this question is to reliably
quantify where on the complexity spectrum a given time series falls; a
second step is to determine how complexity and predictability are
related.  These are the goals of this paper.

@@

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.  
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{KS-entropy}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a periodic
signal, which exhibits low entropy, is on the low end of the
complexity spectrum.  We argue that an extension of \emph{permutation
  entropy}~\cite{bandt2002per}---a method for approximating the
entropy through ordinal analysis---is an effective way to assess the
complexity of a given time series.  Permutation entropy is ideal for
our purposes because it works with real-valued data and is known to
converge to the true entropy value. Other existing techniques either
require specific knowledge of the generating process or produce biased
values of the entropy~\cite{bollt2001}.

@@

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  While this is not a common laboratory
experiment, it is a highly appropriate choice here.  Computers are
among the most complex engineered artifacts in current use.  Modern
microprocessor chips contain multiple processing units and multi-layer
memories, for instance, and they use complicated hardware/software
strategies to move data and threads of computation across those
resources.  As a result, the processor and memory loads during the
execution of even a very simple program can evolve in a very
complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance trace
of a three-line loop that repeatedly initializes the upper triangle of
a matrix in column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic~\cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the program can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate the
  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh
